An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2021-10-14 ju-hofmann/ju-hofmann.tex

\abstract{%
In this paper, we introduce the R package cmpsR, an open-source
implementation of the Congruent Matching Profile Segments (CMPS) method
developed at the National Institute of Standards and Technology (NIST)
for objective comparison of striated tool marks. The functionality of
the package is showcased by examples of bullet signatures that come with
the package. Graphing tools are implemented in the package as well for
users to assess and understand the CMPS results. Initial tests were
performed on bullet signatures generated from two sets of 3D scans in
the Hamby study under the framework suggested by the R package
bulletxtrctr. New metrics based on CMPS scores are introduced and
compared with existing metrics. A measure called Sum of Squares Ratio is
included, and how it can be used for evaluating different scans,
metrics, or parameters is showcased with the Hamby study data sets. An
open-source implementation of the CMPS algorithm makes the algorithm
more accessible, generates reproducible results, and facilitates further
studies of the algorithm such as methods comparison.
}

@@

\textbf{Hamby Data Set}

@@

The automated framework used in this paper was proposed by \citet{aoas}.
In order to obtain bullet signatures from the x3p files of bullet land
engravings, we use the \texttt{R} packages \texttt{x3ptools}
\citep{x3ptools} and \texttt{bulletxtrctr} \citep{bulletxtrctr}.
\texttt{x3ptools} is a package to read, write, and generally, process
x3p files. The \texttt{bulletxtrctr} package implements a pipeline for
extracting and comparing signatures from scans of land-engraved areas.

@@

The white horizontal line in \autoref{fig:process}(a) indicates which
cross-section was identified by the algorithm to represent the LEA;
\autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the
vertical blue lines in \autoref{fig:process}(c) and
\autoref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess}
is fitted to capture the bullet curvature (\autoref{fig:process}(e))
and, finally, the \textbf{bullet signature} (\autoref{fig:process}(f))
is obtained as residuals of the cross-section and the fitted smooth.
Note, that in \citet{cmps} bullet signatures are referred to as bullet
profiles. However, to avoid confusion, we distinguish the notion of
bullet signatures from bullet profiles. Bullet profiles are shown in
panels (b), (c), and (d) of \autoref{fig:process}, while
\autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a
correct down-stream analysis of the signatures. To allow for a
human-in-the middle inspection and intervention we have provided an
interactive web-application, implemented as \texttt{R} Shiny App
\citep{shiny}, \texttt{bulletinspectR} to identify and correct those
errors. An example of the extraction process with corresponding code and
parameter settings can be found on Github at
\url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}.
Note that the process of extracting signatures might be different from
the one used in \citet{cmps} because no code or parameter settings are
made available publicly.

@@

In this example, the comparison between \texttt{sig1} and \texttt{sig2},
two signatures coming from the same source (a known-match comparison),
gets a CMPS score of 17; the comparison between \texttt{sig1} and
\texttt{sig3}, two signatures coming from different sources (a known
non-match comparison), gets a CMPS score of 1.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{cut the comparison signature into consecutive, non-overlapping
  and equal-length basis segments:} The command
  \texttt{get\_segs(x,\ len=50)} implements this step: it takes bullet
  signature \texttt{x} in the format of a numeric vector and cuts it
  into consecutive, non-overlapping and equal-length segments of length
  \texttt{len}, which are referred to as ``basis segments''.
\item
  \textbf{identify candidate positions:} For each basis segment a set of
  candidate registration positions on the comparison signature is
  identified based on the segment's similarity to the reference
  signature. In a first step, the cross-correlation function of the
  segment to the reference is calculated, then a number of positions
  with high correlation values are identified as candidate positions. In
  case multiple segment lengths are considered, the length of each basis
  segment is expanded (by default it is doubled) and these two steps are
  repeated. Only when candidate positions coincide (or are similar
  enough), they are considered further. \autoref{fig:segplots} and
  \autoref{fig:seg_all} illustrate these ideas.

  \begin{itemize}
  \item
    \textbf{Calculate the cross-correlation curve:} Calculate the
    cross-correlation curve between a basis segment \texttt{x} and the
    reference signature \texttt{y} using the function
    \texttt{get\_ccf4(x,\ y,\ ...)} as shown in \autoref{fig:segplot2}.
    The position indicates the lag by which a basis segment is moved
    with respect to its original placement. A position is considered
    ``good'' if it results in a peak in the cross-correlation between
    the basis segment and the reference.
  \item
    \textbf{Correlation peaks:} Two strategies referred to as
    ``multi-peak inspection'' and ``multi-peak inspection at different
    segment lengths'' in \citet{cmps} are used for identifying positions
    of correlation peaks as candidate positions. The latter is also
    called ``multi-segment lengths strategy''. The parameter
    \texttt{npeaks.set} in \texttt{extract\_feature\_cmps(...)}
    determines which strategy to use and the number of candidate
    positions:

    \begin{itemize}
    \tightlist
    \item
      If \texttt{npeaks.set} is an integer vector of length 1, for
      example \texttt{npeaks.set\ =\ 5}, the positions of the top five
      peaks in the cross-correlation curve are identified as candidate
      positions for registration.
    \item
      If \texttt{npeaks.set} is an integer vector of length more than 1,
      for example \texttt{npeaks.set\ =\ c(5,\ 3,\ 1)}, the
      multi-segment lengths strategy will be used: calculate the
      cross-correlation function between a basis segment and the
      reference and identify positions of the top five peaks; double the
      segment length to a specified value, re-calculate the
      cross-correlation function, and identify three peaks; repeat this
      process and identify a single peak in the newly computed
      cross-correlation function. \autoref{fig:seg_all} shows an example
      of three levels of basis segment 6 and their corresponding
      cross-correlation curves and identified peaks. Note that in
      \citet{cmps} the segment length is doubled at each level of a
      basis segment, but in the present implementation users are allowed
      to choose the segment length at each level.
    \item
      \texttt{get\_ccr\_peaks(comp,\ segments,\ seg\_outlength,\ nseg\ =\ 1,\ npeaks\ =\ 5)}
      computes the cross-correlation curve between a basis or increased
      segment and the reference signature and finds peaks in the
      cross-correlation curve. The number of peaks detected is equal to
      \texttt{npeaks}, which is an integer. \texttt{segments},
      \texttt{seg\_outlength}, and \texttt{nseg} determines the segment
      in the cross-correlation computation, and \texttt{comp} gives the
      reference signature. If the multi-segment lengths strategy is
      used, then \texttt{get\_ccr\_peaks(...)} is called in a
      \texttt{lapply()} for each level of the basis segment. The
      resulting list is called \texttt{ccr.list}.
    \end{itemize}
  \item
    \textbf{multi-segment lengths strategy:} with the multi-segment
    lengths strategy being used, a position is identified as a candidate
    position for registration and is called as a ``consistent
    correlation peak'' if it results in a top peak in the
    cross-correlation curve with a tolerance zone determined by
    \texttt{Tx} in all segment levels. Note that in \citet{cmps}, a
    segment at its largest scale (highest level) always identifies one
    peak, but we do not have this requirement in our implementation.

    \begin{itemize}
    \tightlist
    \item
      the function
      \texttt{get\_seg\_scale(segments,\ nseg,\ out\_length)} is used to
      obtain the (potentially increased) version of a basis segment.
      \texttt{segments}, which is a list containing all basis segments
      generated by the function \texttt{get\_segs(...)} in step 1, and
      \texttt{nseg} are used to determine the basis segment to be
      increased. \texttt{out\_length} specifies the length of the output
      segment.
    \item
      \texttt{get\_ccp(ccr.list,\ Tx\ =\ 25)} tries to identify the
      ``consistent correlation peak''. \texttt{ccr.list} is the result
      of \texttt{lapply()} and \texttt{get\_ccr\_peaks(...)}, and
      \texttt{Tx} determines the size of a tolerance zone used in
      identifying the consistent correlation peak.
      \texttt{get\_ccp(...)} returns \texttt{NULL} if there is no
      consistent correlation peak.
    \end{itemize}
  \end{itemize}
\item
  \textbf{determine the congruent registration position:} A candidate
  position ``receives'' \texttt{votes} from basis segments that identify
  it or a close position within a tolerance zone of \texttt{Tx} as a
  candidate position in step 2. Votes for all candidate positions are
  tallied, and the position with the highest number of votes gets chosen
  as the \emph{congruent registration position}, indicating that most of
  the basis segments find their highly similar counterpart in the
  reference signature in terms of correlation at this registration
  position. In case of ties the middle position is taken as the
  congruent registration position. Basis segments with a congruent
  registration position are called ``congruent matching profile
  segments'' (CMPS). The total number of CMPS is the CMPS score of the
  comparison. \texttt{get\_CMPS(input.ccp,\ Tx\ =\ 25)} is the function
  that tallies the votes and determines the congruent registration
  position and congruent matching profile segments (CMPS).
\end{enumerate}

Note that there are several parameters in the CMPS algorithm that are
left to the users to decide, such as the length of basis segments
\texttt{seg\_length} in step 1, the number of peaks \texttt{npeaks.set}
identified on each level in step 2, and the length of the tolerance zone
\texttt{Tx} in both step 2 and 3. In our implementation of the CMPS
algorithm we used the parameters given in the original CMPS paper
\citep{cmps} as the default values for these parameters. However, the
authors state that no cross-validation has been done - there might be
also issues with respect to resolution of the scans. Further research is
needed, until then users are advised to think of default values as
starting values and consider alternatives. The main function that
combines all steps in the CMPS algorithm described above is called
\texttt{extract\_feature\_cmps(...)}. Here we present it with its
default parameters.

@@

The function \texttt{extract\_feature\_cmps} allows for the following
input from users besides the previously discussed parameters
\texttt{seg\_length}, \texttt{npeaks.set}, and \texttt{Tx}:

\begin{itemize}
\item
  \texttt{x} and \texttt{y} are two signatures: \texttt{x} serves as the
  comparison signature (which will be divided into basis segments) and
  \texttt{y} is the reference signature;
\item
  \texttt{include} determines the format of the function result. Besides
  the \texttt{CMPS.score}, other aspects of the comparison help in
  understanding how the \texttt{CMPS.score} is computed. By default
  \texttt{include} is set to \texttt{NULL} and only the
  \texttt{CMPS.score} is returned; further results are included when
  \texttt{include} is (an abbreviation of) one of or a vector of the
  following strings:
  \texttt{"nseg",\ "congruent.pos",\ "congruent.seg.idx",\ "segments",\ "parameters"},
  and \texttt{"full\_result"}. If \texttt{include} is specified as
  \texttt{"full\_result"} (or its abbreviation), the output includes
  everything listed below.

  \begin{itemize}
  \tightlist
  \item
    \texttt{nseg}: the number of basis segments from the comparison
    signature; this is also the highest possible CMPS score of the
    comparison;
  \item
    \texttt{congruent.pos}: the congruent registration position;
  \item
    \texttt{congruent.seg.idx}: the indices of all congruent matching
    profile segments;
  \item
    \texttt{ccp.list}: a list showing identified candidate positions of
    all basis segments;
  \item
    \texttt{pos.df}: a data frame containing all candidate positions and
    their respective number of votes;
  \item
    \texttt{segments}: a list containing all basis segments;
  \item
    \texttt{parameters}: a list containing all input arguments of
    \texttt{extract\_feature\_cmps};
  \end{itemize}
\item
  \texttt{outlength} specifies the segment length of a basis segment at
  each level under the multi-segment lengths strategy. By default
  \texttt{outlength} is set to \texttt{NULL}, indicating that a basis
  segment should double its segment length at the next level and
  conforming to the description in \citet{cmps} .
\end{itemize}

@@

\textbf{An Example}

@@

\begin{Schunk}
\begin{Sinput}
library(cmpsR)
\end{Sinput}
\begin{Soutput}
#> 
#> Attaching package: 'cmpsR'
\end{Soutput}
\begin{Soutput}
#> The following objects are masked _by_ '.GlobalEnv':
#> 
#>     compute_diff_phase, compute_score_metrics,
#>     get_all_phases, metric_plot_helper
\end{Soutput}
\begin{Sinput}
data(bullets)
\end{Sinput}
\end{Schunk}

\texttt{bullets\$sigs} is a list of twelve numeric vectors corresponding
to the twelve bullet signatures shown in \autoref{fig:sigs}.
\texttt{bullets\$source} contains the URLs to the corresponding x3p file
containing the topographic scan from the NIST Ballistics Toolmark
Research Database \citep{nistdb}.

@@

The signatures of Land 4 of Bullet 1 and Land 5 of Bullet 2 are stored
in objects \texttt{sigs2} and \texttt{sigs1}, respectively. This
comparison consists of a pair of signatures that are known to be a match
-- a KM (known match) comparison. We compute the CMPS score using two
versions of the CMPS algorithm:

@@

In the first example, \texttt{npeaks.set} is a vector of three integers,
i.e.~the algorithm uses the multi-segment lengths strategy to create the
result object \texttt{cmps\_with\_multi\_scale}. For
\texttt{cmps\_without\_multi\_scale} each basis segment is linked to the
top 5 candidate positions. We use \texttt{include\ =\ "full\_result"} to
capture all results. In this example, the CMPS is 9 when using multiple
segments, and 12 when using a single segment. As discussed in
\citet{cmps}, using multi-segment lengths strategy can reduce the number
of false positives when identifying candidate positions; however, any
score based method is walking the line between false positives and false
negatives. As the number of false positives is reduced the number of
false negatives might rise. More discussion and comparisons between the
two versions of the CMPS algorithm will be presented in later sections.
Note, that the multi-segment lengths method is slower because the
algorithm is run once for each segment length.

\textbf{Visualize and Understand CMPS results}

We also implemented graphing tools for visualizing results of the CMPS
algorithm. The goal is to provide users with tools to inspect each of
the basis segments and to help them have a better understanding of how
the algorithm works. \autoref{fig:sigplots} shows the plots generated by
the first graphing function, \texttt{cmps\_signature\_plot()}, and
continues with the example above. \texttt{cmps\_signature\_plot()} takes
the output of
\texttt{extract\_feature\_cmps(...,\ include\ =\ "full\_result")} and
returns a list of 5 elements. It creates an overall impression of how
the comparison signature aligns with the reference signature at the
congruent registration position.

\begin{itemize}
\item
  The first element is a plot called \texttt{segment\_shift\_plot},
  shown in \autoref{fig:sigplot}. On this plot the reference signature
  is drawn as a black line, congruent matching profile segments from the
  comparison signature are overlaid in red at the congruent registration
  position.
\item
  The second plot is called \texttt{signature\_shift\_plot}, shown in
  \autoref{fig:sigplot2}. This visual presents both the comparison
  signature and the reference signature. The comparison signature is
  aligned with the reference signature based on the congruent
  registration position. Congruent matching profile segments are
  highlighted by solid red lines.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Other elements of this list are \texttt{seg\_shift} and
  \texttt{sig\_shift}. \texttt{sig\_shift} gives the congruent
  registration position, while \texttt{seg\_shift} is a data frame
  showing the congruent matching profile segments and their identified
  candidate position closest to the congruent registration position.
\end{itemize}

@@

While \texttt{cmps\_signature\_plot()} focuses on the signature level,
\texttt{cmps\_segment\_plot()} focuses on the segment level. It provides
the ``full result'' of \texttt{extract\_feature\_cmps()}, but also takes
an argument, \texttt{seg.idx}, indicating which segment should be
inspected. When checking \texttt{sig.plot\$seg\_shift} we notice that
segment number 6 is not one of the congruent matching profile segments.
We can therefore set \texttt{seg.idx\ =\ 6} in
\texttt{cmps\_segment\_plot()} and investigate the reason why this
segment disagrees with the congruent registration position.

For each segment scale, we have two plots: \texttt{segment\_plot} and
\texttt{scale\_ccf\_plot}, as shown in \autoref{fig:segplots} for the
example of segment number 6:

\begin{itemize}
\tightlist
\item
  \autoref{fig:segplot} is the \texttt{segment\_plot} for basis segment
  6 at level one (in its original length). We used
  \texttt{npeaks.set\ =\ c(5,\ 3,\ 1)} in
  \texttt{extract\_feature\_cmps()} when calculating the CMPS score.
  Therefore the top five peaks are identified in the cross-correlation
  curve at level one. Segment 6 is plotted at the positions where these
  5 peaks are identified with dashed lines in the
  \texttt{segment\_plot}. The solid thick black line shows the segment
  at its original position (which in this example is very close to the
  actual registration position).
\item
  \autoref{fig:segplot2} is the \texttt{scale\_ccf\_plot} of basis
  segment 6 at level one. It shows the cross-correlation curve computed
  by the reference signature and the level-one basis segment 6. The five
  highest peaks are marked by dots on the curve.
\end{itemize}

@@

Additionally, users can have more insights about why segment 6 is not a
congruent matching profile segment if we put the \texttt{segment\_plot}
and \texttt{scale\_ccf\_plot} of all three segment levels together, as
shown in \autoref{fig:seg_all} with the help of
\texttt{ggpubr::ggarrange()}.

@@

\hypertarget{evaluation-metrics}{%
\subsection{Evaluation Metrics}\label{evaluation-metrics}}

\textbf{Metrics Based on CMPS scores}

@@

We can continue with the example used in previous sections.
\texttt{bullets} contains bullet signatures of two bullets,
\texttt{bullet1} and \texttt{bullet2}. As mentioned before, each bullet
has 6 land engravings, resulting in 6 bullet signatures. Thus, there are
36 pairwise bullet signature comparisons, resulting in 36 \(c_{ij}\)
values in total. We use multi-segment lengths strategy with default
parameters to compute these CMPS scores, and the result is shown in
\autoref{fig:tiles}. We can see that in this example,

@@

\textbf{Scaled CMPS Scores}

@@

\textbf{Sum of Squares Ratio}

@@

\autoref{fig:result1_252} shows the distribution of
\(\mathrm{CMPS_{max}}\) and \(\mathrm{\overline{CMPS}_{max}}\) after we
applied the CMPS algorithm to Hamby set 252 with the multi-segment
lengths strategy. The parameters we used in
\texttt{extract\_feature\_cmps} for the CMPS algorithm are:

@@

Additionally, to mimic the parameters used in \citet{cmps}, we set
\texttt{seg\_length\ =\ 50} and \texttt{Tx\ =\ 25} to make sure that
each basis segment has length of 78.125 \textmu m and the tolerance zone
is \(\pm 39.0625\) \textmu m (one unit represents 1.5625 \textmu m for
Hamby set 252).

>>>>

\abstract{%
In this paper, we introduce the R package \pkg{cmpsR}, an open-source
implementation of the Congruent Matching Profile Segments (CMPS) method
developed at the National Institute of Standards and Technology (NIST)
for objective comparison of striated tool marks. The functionality of
the package is showcased by examples of bullet signatures that come with
the package. Graphing tools are implemented in the package as well for
users to assess and understand the CMPS results. Initial tests were
performed on bullet signatures generated from two sets of 3D scans in
the Hamby study under the framework suggested by the R package
\pkg{bulletxtrctr}. New metrics based on CMPS scores are introduced and
compared with existing metrics. A measure called Sum of squares ratio is
included, and how it can be used for evaluating different scans,
metrics, or parameters is showcased with the Hamby study data sets. An
open-source implementation of the CMPS algorithm makes the algorithm
more accessible, generates reproducible results, and facilitates further
studies of the algorithm such as methods comparison.
}

@@

\textbf{Hamby data set}

@@

The automated framework used in this paper was proposed by \citet{aoas}.
In order to obtain bullet signatures from the x3p files of bullet land
engravings, we use the \texttt{R} packages \pkg{x3ptools}
\citep{x3ptools} and \pkg{bulletxtrctr} \citep{bulletxtrctr}.
\pkg{x3ptools} is a package to read, write, and generally, process x3p
files. The \pkg{bulletxtrctr} package implements a pipeline for
extracting and comparing signatures from scans of land-engraved areas.

@@

The white horizontal line in \autoref{fig:process}(a) indicates which
cross-section was identified by the algorithm to represent the LEA;
\autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the
vertical blue lines in \autoref{fig:process}(c) and
\autoref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess}
is fitted to capture the bullet curvature (\autoref{fig:process}(e))
and, finally, the \textbf{bullet signature} (\autoref{fig:process}(f))
is obtained as residuals of the cross-section and the fitted smooth.
Note, that in \citet{cmps} bullet signatures are referred to as bullet
profiles. However, to avoid confusion, we distinguish the notion of
bullet signatures from bullet profiles. Bullet profiles are shown in
panels (b), (c), and (d) of \autoref{fig:process}, while
\autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a
correct down-stream analysis of the signatures. To allow for a
human-in-the middle inspection and intervention we have provided an
interactive web-application, implemented as \texttt{R} Shiny App
\citep{shiny}, \pkg{bulletinspectR} to identify and correct those
errors. An example of the extraction process with corresponding code and
parameter settings can be found on Github at
\url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}.
Note that the process of extracting signatures might be different from
the one used in \citet{cmps} because no code or parameter settings are
made available publicly.

@@

In this example, the comparison between \code{sig1} and \code{sig2}, two
signatures coming from the same source (a known-match comparison), gets
a CMPS score of 17; the comparison between \code{sig1} and \code{sig3},
two signatures coming from different sources (a known non-match
comparison), gets a CMPS score of 1.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{cut the comparison signature into consecutive, non-overlapping
  and equal-length basis segments:} The command
  \code{get\_segs(x, len=50)} implements this step: it takes bullet
  signature \code{x} in the format of a numeric vector and cuts it into
  consecutive, non-overlapping and equal-length segments of length
  \code{len}, which are referred to as ``basis segments''.
\item
  \textbf{identify candidate positions:} For each basis segment a set of
  candidate registration positions on the comparison signature is
  identified based on the segment's similarity to the reference
  signature. In a first step, the cross-correlation function of the
  segment to the reference is calculated, then a number of positions
  with high correlation values are identified as candidate positions. In
  case multiple segment lengths are considered, the length of each basis
  segment is expanded (by default it is doubled) and these two steps are
  repeated. Only when candidate positions coincide (or are similar
  enough), they are considered further. \autoref{fig:segplots} and
  \autoref{fig:seg_all} illustrate these ideas.

  \begin{itemize}
  \item
    \textbf{Calculate the cross-correlation curve:} Calculate the
    cross-correlation curve between a basis segment \code{x} and the
    reference signature \code{y} using the function
    \code{get\_ccf4(x, y, ...)} as shown in \autoref{fig:segplot2}. The
    position indicates the lag by which a basis segment is moved with
    respect to its original placement. A position is considered ``good''
    if it results in a peak in the cross-correlation between the basis
    segment and the reference.
  \item
    \textbf{Correlation peaks:} Two strategies referred to as
    ``multi-peak inspection'' and ``multi-peak inspection at different
    segment lengths'' in \citet{cmps} are used for identifying positions
    of correlation peaks as candidate positions. The latter is also
    called ``multi-segment lengths strategy''. The parameter
    \code{npeaks.set} in \code{extract\_feature\_cmps(...)} determines
    which strategy to use and the number of candidate positions:

    \begin{itemize}
    \tightlist
    \item
      If \code{npeaks.set} is an integer vector of length 1, for example
      \code{npeaks.set = 5}, the positions of the top five peaks in the
      cross-correlation curve are identified as candidate positions for
      registration.
    \item
      If \code{npeaks.set} is an integer vector of length more than 1,
      for example \code{npeaks.set = c(5, 3, 1)}, the multi-segment
      lengths strategy will be used: calculate the cross-correlation
      function between a basis segment and the reference and identify
      positions of the top five peaks; double the segment length to a
      specified value, re-calculate the cross-correlation function, and
      identify three peaks; repeat this process and identify a single
      peak in the newly computed cross-correlation function.
      \autoref{fig:seg_all} shows an example of three levels of basis
      segment 6 and their corresponding cross-correlation curves and
      identified peaks. Note that in \citet{cmps} the segment length is
      doubled at each level of a basis segment, but in the present
      implementation users are allowed to choose the segment length at
      each level.
    \item
      \code{get\_ccr\_peaks(comp, segments, seg\_outlength, nseg = 1, npeaks = 5)}
      computes the cross-correlation curve between a basis or increased
      segment and the reference signature and finds peaks in the
      cross-correlation curve. The number of peaks detected is equal to
      \code{npeaks}, which is an integer. \code{segments},
      \code{seg\_outlength}, and \code{nseg} determines the segment in
      the cross-correlation computation, and \code{comp} gives the
      reference signature. If the multi-segment lengths strategy is
      used, then \code{get\_ccr\_peaks(...)} is called in a
      \code{lapply()} for each level of the basis segment. The resulting
      list is called \code{ccr.list}.
    \end{itemize}
  \item
    \textbf{multi-segment lengths strategy:} with the multi-segment
    lengths strategy being used, a position is identified as a candidate
    position for registration and is called as a ``consistent
    correlation peak'' if it results in a top peak in the
    cross-correlation curve with a tolerance zone determined by
    \code{Tx} in all segment levels. Note that in \citet{cmps}, a
    segment at its largest scale (highest level) always identifies one
    peak, but we do not have this requirement in our implementation.

    \begin{itemize}
    \tightlist
    \item
      the function \code{get\_seg\_scale(segments, nseg, out\_length)}
      is used to obtain the (potentially increased) version of a basis
      segment. \code{segments}, which is a list containing all basis
      segments generated by the function \code{get\_segs(...)} in step
      1, and \code{nseg} are used to determine the basis segment to be
      increased. \code{out\_length} specifies the length of the output
      segment.
    \item
      \code{get\_ccp(ccr.list, Tx = 25)} tries to identify the
      ``consistent correlation peak''. \code{ccr.list} is the result of
      \code{lapply()} and \code{get\_ccr\_peaks(...)}, and \code{Tx}
      determines the size of a tolerance zone used in identifying the
      consistent correlation peak. \code{get\_ccp(...)} returns
      \code{NULL} if there is no consistent correlation peak.
    \end{itemize}
  \end{itemize}
\item
  \textbf{determine the congruent registration position:} A candidate
  position ``receives'' votes from basis segments that identify it or a
  close position within a tolerance zone of \code{Tx} as a candidate
  position in step 2. Votes for all candidate positions are tallied, and
  the position with the highest number of votes gets chosen as the
  \emph{congruent registration position}, indicating that most of the
  basis segments find their highly similar counterpart in the reference
  signature in terms of correlation at this registration position. In
  case of ties the middle position is taken as the congruent
  registration position. Basis segments with a congruent registration
  position are called ``congruent matching profile segments'' (CMPS).
  The total number of CMPS is the CMPS score of the comparison.
  \code{get\_CMPS(input.ccp, Tx = 25)} is the function that tallies the
  votes and determines the congruent registration position and congruent
  matching profile segments (CMPS).
\end{enumerate}

Note that there are several parameters in the CMPS algorithm that are
left to the users to decide, such as the length of basis segments
\code{seg\_length} in step 1, the number of peaks \code{npeaks.set}
identified on each level in step 2, and the length of the tolerance zone
\code{Tx} in both step 2 and 3. In our implementation of the CMPS
algorithm we used the parameters given in the original CMPS paper
\citep{cmps} as the default values for these parameters. However, the
authors state that no cross-validation has been done - there might be
also issues with respect to resolution of the scans. Further research is
needed, until then users are advised to think of default values as
starting values and consider alternatives. The main function that
combines all steps in the CMPS algorithm described above is called
\code{extract\_feature\_cmps(...)}. Here we present it with its default
parameters.

@@

The function \code{extract\_feature\_cmps} allows for the following
input from users besides the previously discussed parameters
\code{seg\_length}, \code{npeaks.set}, and \code{Tx}:

\begin{itemize}
\item
  \code{x} and \code{y} are two signatures: \code{x} serves as the
  comparison signature (which will be divided into basis segments) and
  \code{y} is the reference signature;
\item
  \code{include} determines the format of the function result. Besides
  the \code{CMPS.score}, other aspects of the comparison help in
  understanding how the \code{CMPS.score} is computed. By default
  \code{include} is set to \code{NULL} and only the \code{CMPS.score} is
  returned; further results are included when \code{include} is (an
  abbreviation of) one of or a vector of the following strings:
  \code{"nseg"}, \code{"congruent.pos"}, \code{"congruent.seg.idx"},
  \code{"segments"}, \code{"parameters"}, and \code{"full\_result"}. If
  \code{include} is specified as \code{"full\_result"} (or its
  abbreviation), the output includes everything listed below.

  \begin{itemize}
  \tightlist
  \item
    \code{nseg}: the number of basis segments from the comparison
    signature; this is also the highest possible CMPS score of the
    comparison;
  \item
    \code{congruent.pos}: the congruent registration position;
  \item
    \code{congruent.seg.idx}: the indices of all congruent matching
    profile segments;
  \item
    \code{ccp.list}: a list showing identified candidate positions of
    all basis segments;
  \item
    \code{pos.df}: a data frame containing all candidate positions and
    their respective number of votes;
  \item
    \code{segments}: a list containing all basis segments;
  \item
    \code{parameters}: a list containing all input arguments of
    \code{extract\_feature\_cmps};
  \end{itemize}
\item
  \code{outlength} specifies the segment length of a basis segment at
  each level under the multi-segment lengths strategy. By default
  \code{outlength} is set to \code{NULL}, indicating that a basis
  segment should double its segment length at the next level and
  conforming to the description in \citet{cmps} .
\end{itemize}

@@

\textbf{An example}

@@

\begin{Schunk}
\begin{Sinput}
library(cmpsR)
data(bullets)
\end{Sinput}
\end{Schunk}

\code{bullets\$sigs} is a list of twelve numeric vectors corresponding
to the twelve bullet signatures shown in \autoref{fig:sigs}.
\code{bullets\$source} contains the URLs to the corresponding x3p file
containing the topographic scan from the NIST Ballistics Toolmark
Research Database \citep{nistdb}.

@@

The signatures of Land 4 of Bullet 1 and Land 5 of Bullet 2 are stored
in objects \code{sigs2} and \code{sigs1}, respectively. This comparison
consists of a pair of signatures that are known to be a match -- a KM
(known match) comparison. We compute the CMPS score using two versions
of the CMPS algorithm:

@@

In the first example, \code{npeaks.set} is a vector of three integers,
i.e.~the algorithm uses the multi-segment lengths strategy to create the
result object \code{cmps\_with\_multi\_scale}. For
\code{cmps\_without\_multi\_scale} each basis segment is linked to the
top 5 candidate positions. We use \code{include = "full\_result"} to
capture all results. In this example, the CMPS is 9 when using multiple
segments, and 12 when using a single segment. As discussed in
\citet{cmps}, using multi-segment lengths strategy can reduce the number
of false positives when identifying candidate positions; however, any
score based method is walking the line between false positives and false
negatives. As the number of false positives is reduced the number of
false negatives might rise. More discussion and comparisons between the
two versions of the CMPS algorithm will be presented in later sections.
Note, that the multi-segment lengths method is slower because the
algorithm is run once for each segment length.

\textbf{Visualize and understand CMPS results}

We also implemented graphing tools for visualizing results of the CMPS
algorithm. The goal is to provide users with tools to inspect each of
the basis segments and to help them have a better understanding of how
the algorithm works. \autoref{fig:sigplots} shows the plots generated by
the first graphing function, \code{cmps\_signature\_plot()}, and
continues with the example above. \code{cmps\_signature\_plot()} takes
the output of
\code{extract\_feature\_cmps(..., include = "full\_result")} and returns
a list of 5 elements. It creates an overall impression of how the
comparison signature aligns with the reference signature at the
congruent registration position.

\begin{itemize}
\item
  The first element is a plot called \code{segment\_shift\_plot}, shown
  in \autoref{fig:sigplot}. On this plot the reference signature is
  drawn as a black line, congruent matching profile segments from the
  comparison signature are overlaid in red at the congruent registration
  position.
\item
  The second plot is called \code{signature\_shift\_plot}, shown in
  \autoref{fig:sigplot2}. This visual presents both the comparison
  signature and the reference signature. The comparison signature is
  aligned with the reference signature based on the congruent
  registration position. Congruent matching profile segments are
  highlighted by solid red lines.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Other elements of this list are \code{seg\_shift} and
  \code{sig\_shift}. \code{sig\_shift} gives the congruent registration
  position, while \code{seg\_shift} is a data frame showing the
  congruent matching profile segments and their identified candidate
  position closest to the congruent registration position.
\end{itemize}

@@

While \code{cmps\_signature\_plot()} focuses on the signature level,
\code{cmps\_segment\_plot()} focuses on the segment level. It provides
the ``full result'' of \code{extract\_feature\_cmps()}, but also takes
an argument, \code{seg.idx}, indicating which segment should be
inspected. When checking \code{sig.plot\$seg\_shift} we notice that
segment number 6 is not one of the congruent matching profile segments.
We can therefore set \code{seg.idx = 6} in \code{cmps\_segment\_plot()}
and investigate the reason why this segment disagrees with the congruent
registration position.

For each segment scale, we have two plots: \code{segment\_plot} and
\code{scale\_ccf\_plot}, as shown in \autoref{fig:segplots} for the
example of segment number 6:

\begin{itemize}
\tightlist
\item
  \autoref{fig:segplot} is the \code{segment\_plot} for basis segment 6
  at level one (in its original length). We used
  \code{npeaks.set = c(5, 3, 1)} in \code{extract\_feature\_cmps()} when
  calculating the CMPS score. Therefore the top five peaks are
  identified in the cross-correlation curve at level one. Segment 6 is
  plotted at the positions where these 5 peaks are identified with
  dashed lines in the \code{segment\_plot}. The solid thick black line
  shows the segment at its original position (which in this example is
  very close to the actual registration position).
\item
  \autoref{fig:segplot2} is the \code{scale\_ccf\_plot} of basis segment
  6 at level one. It shows the cross-correlation curve computed by the
  reference signature and the level-one basis segment 6. The five
  highest peaks are marked by dots on the curve.
\end{itemize}

@@

Additionally, users can have more insights about why segment 6 is not a
congruent matching profile segment if we put the \code{segment\_plot}
and \code{scale\_ccf\_plot} of all three segment levels together, as
shown in \autoref{fig:seg_all} with the help of
\code{ggpubr::ggarrange()}.

@@

\hypertarget{evaluation-metrics}{%
\subsection{Evaluation metrics}\label{evaluation-metrics}}

\textbf{Metrics based on CMPS scores}

@@

We can continue with the example used in previous sections.
\code{bullets} contains bullet signatures of two bullets, \code{bullet1}
and \code{bullet2}. As mentioned before, each bullet has 6 land
engravings, resulting in 6 bullet signatures. Thus, there are 36
pairwise bullet signature comparisons, resulting in 36 \(c_{ij}\) values
in total. We use multi-segment lengths strategy with default parameters
to compute these CMPS scores, and the result is shown in
\autoref{fig:tiles}. We can see that in this example,

@@

\textbf{Scaled CMPS scores}

@@

\textbf{Sum of squares ratio}

@@

\autoref{fig:result1_252} shows the distribution of
\(\mathrm{CMPS_{max}}\) and \(\mathrm{\overline{CMPS}_{max}}\) after we
applied the CMPS algorithm to Hamby set 252 with the multi-segment
lengths strategy. The parameters we used in
\code{extract\_feature\_cmps} for the CMPS algorithm are:

@@

Additionally, to mimic the parameters used in \citet{cmps}, we set
\code{seg\_length = 50} and \code{Tx = 25} to make sure that each basis
segment has length of 78.125 \textmu m and the tolerance zone is
\(\pm 39.0625\) \textmu m (one unit represents 1.5625 \textmu m for
Hamby set 252).
